---
output: github_document
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# optimflex

`optimflex` provides a highly flexible suite of derivative-based non-linear optimization algorithms. It is specifically designed for researchers who require **rigorous convergence control**, particularly in complex models like SEM.

## Why Use optimflex?

Standard optimization functions often rely on a single, fixed stopping rule. `optimflex` offers:

1. **Strict Convergence Control**: Choose from 8 distinct criteria.
2. **The "AND" Rule**: All selected criteria must be met simultaneously.
3. **Hessian Verification**: Verify local minima by checking positive definiteness at the final point.

## Installation

```r
# install.packages("devtools")
# devtools::install_github("yourusername/optimflex")
```

## Basic Usage

```{r example}
library(optimflex)

rosenbrock <- function(x) {
  100 * (x[2] - x[1]^2)^2 + (1 - x[1])^2
}

res <- double_dogleg(
  start = c(-1.2, 1.0),
  objective = rosenbrock,
  control = list(
    use_grad = TRUE,
    use_rel_x = TRUE,
    use_posdef = TRUE
  )
)

print(res$par)
```

## Convergence Criteria Overview

| Flag | Description |
|:---|:---|
| `use_abs_f` | Absolute function change |
| `use_rel_f` | Relative function change |
| `use_abs_x` | Absolute parameter change |
| `use_rel_x` | Relative parameter change |
| `use_grad`  | Gradient infinity norm |
| `use_posdef`| Hessian Verification |
| `use_pred_f`| Predicted Decrease |
